Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-09-10 Manuscript.qmd
testing
Minor edits to manuscript, fiddling with pdf options

bibliography: references.bib
csl: apa.csl 
embed-resources: true
---

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy a difficult quantity to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data. Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017]. Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

@@

modellingSummary |>
  mutate(Variable = fct(Variable, 
                        levels = c("Number", "MedianCovs", 
                                   "Prop_AnySelect", "Prop_Candidate",
                                   "Prop_Sequential", "Prop_Simple",
                                   "Prop_ModAvg", "Prop_GOF", 
                                   "Prop_Eval"))) |>
  arrange(Variable) |>
  mutate(Variable = as.character(Variable)) |>
  gt(rowname_col = "Variable") |> 
  tab_row_group(
    label = "Covariate selection methods",
    rows = c("Prop_AnySelect", "Prop_Candidate",
             "Prop_Sequential", "Prop_Simple", "Prop_ModAvg"),
    id = "select"
  ) |>
  tab_row_group(
    label = "Model evaluation conducted",
    rows = c("Prop_GOF", "Prop_Eval"),
    id = "eval"
  ) |>
  row_group_order(groups = c(NA, "select", "eval")) |>
  text_case_match("Number" ~ "Number of studies",
                  "MedianCovs" ~ "Median covariates considered per parameter",
                  "Prop_ModAvg" ~ "Percentage using model-averaging",
                  "Prop_AnySelect" ~ "Percentage using any model selection approach",
                  "Prop_Sequential" ~ "Percentage using procedural model selection",
                  "Prop_Simple" ~ "Percentage selecting covariates with simpler models",
                  "Prop_Candidate" ~ "Percentage comparing models in a candidate set",
                  "Prop_GOF" ~ "Percentage calculating goodness-of-fit",
                  "Prop_Eval" ~ "Percentage assessing predictive performance",
                  .locations = cells_stub()) |>
  fmt_percent(rows = 3:9, decimals = 0) |>
  fmt_integer(rows = 1) |>
  tab_stub_indent(
    rows = 3:9,
    indent = 3
  ) |>
  tab_style(
    locations = cells_row_groups(),
    style = cell_text(style = "italic")
  ) |>
  tab_style(
    locations = cells_column_labels(),
    style = list(cell_fill(color = "aquamarine4"),
                 cell_text(color = "white"))
  ) |>
  tab_style(
    locations = cells_body(columns = "All models"),
    style = cell_text(style = "italic")
  ) |>
  as_raw_html()
```

>>>>

bibliography: references.bib
csl: apa.csl
---

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy a difficult quantity to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data. Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017]. Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

@@

modellingSummary |>
  mutate(Variable = fct(Variable, 
                        levels = c("Number", "MedianCovs", 
                                   "Prop_AnySelect", "Prop_Candidate",
                                   "Prop_Sequential", "Prop_Simple",
                                   "Prop_ModAvg", "Prop_GOF", 
                                   "Prop_Eval"))) |>
  arrange(Variable) |>
  mutate(Variable = as.character(Variable)) |>
  gt(rowname_col = "Variable") |> 
  tab_row_group(
    label = "Covariate selection methods",
    rows = c("Prop_AnySelect", "Prop_Candidate",
             "Prop_Sequential", "Prop_Simple", "Prop_ModAvg"),
    id = "select"
  ) |>
  tab_row_group(
    label = "Model evaluation conducted",
    rows = c("Prop_GOF", "Prop_Eval"),
    id = "eval"
  ) |>
  row_group_order(groups = c(NA, "select", "eval")) |>
  text_case_match("Number" ~ "Number of studies",
                  "MedianCovs" ~ "Median covariates considered per parameter",
                  "Prop_ModAvg" ~ "Percentage using model-averaging",
                  "Prop_AnySelect" ~ "Percentage using any model selection approach",
                  "Prop_Sequential" ~ "Percentage using procedural model selection",
                  "Prop_Simple" ~ "Percentage selecting covariates with simpler models",
                  "Prop_Candidate" ~ "Percentage comparing models in a candidate set",
                  "Prop_GOF" ~ "Percentage calculating goodness-of-fit",
                  "Prop_Eval" ~ "Percentage assessing predictive performance",
                  .locations = cells_stub()) |>
  fmt_percent(rows = 3:9, decimals = 0) |>
  fmt_integer(rows = 1) |>
  tab_stub_indent(
    rows = 3:9,
    indent = 3
  ) |>
  tab_style(
    locations = cells_row_groups(),
    style = cell_text(style = "italic")
  ) |>
  tab_style(
    locations = cells_column_labels(),
    style = list(cell_fill(color = "aquamarine4"),
                 cell_text(color = "white"))
  ) |>
  tab_style(
    locations = cells_body(columns = "All models"),
    style = cell_text(style = "italic")
  ) 
#|>
 # as_raw_html()
```
